Forecasting the Support of Kamala Harris in 2024 U.S. Presidential Election using a Model-Based Approach

### 2024-11-04 paper/ForecastingtheSupportofKamalaHarrisin2024U.S.PresidentialElection.qmd
update paper with figure caption

ggplot(harris_data, aes(x = end_date, y = pct, color = factor(numeric_grade))) +
  geom_point(alpha = 0.7) +
  geom_smooth(se = FALSE, span = 0.3) +
  labs(y = "Support for Harris (%)", x = "End Date of the Poll", color = "Numeric Grade") +
  scale_color_viridis_d() +
  theme_classic() +
  theme(legend.position = "bottom") +
  coord_cartesian(ylim = c(0, 100))
```
@tbl-harrissupport shows Harris' average support across states in the 2024 U.S. Presidential Election. She leads in Democratic-leaning states, with top support in Vermont (68.5%), Maryland (62.9%), and Massachusetts (61.2%). In key battlegrounds, Harris's support hovers around 50%, including Wisconsin (49.0%) and Pennsylvania (48.3%). Nationally, she averages 47.9%, with competitive support in swing states like Nevada (47.9%) and North Carolina (47.8%). However, in Republican-leaning states like Texas (44.3%) and Florida (44.2%), her support is lower. This distribution highlights Harris’s strength in Democratic regions and mixed prospects in competitive areas.
```{r}
#| label: tbl-harrissupport
#| tbl-cap: Harris leads in Democratic-leaning states, showing strong support in reliably blue areas. In pivotal swing states, she maintains competitive support, with Wisconsin at 49% and Pennsylvania at 48.3%, reflecting her appeal in regions important for a potential electoral victory. However, in traditionally Republican strongholds, indicating areas where additional campaign efforts might bolster her standing. Nationally, Harris averages 47.9%, underscoring the competitive landscape as election day nears.
#| echo: false
#| warning: false
#| message: false

@@

# Display the result as a table
kable(harris_data_state, caption = "Average Support for Kamala Harris by State")

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| fig-cap: "Harris Kamala leads Donald Trump an average of 1% in Predicted Support in the 2024 US Election (From August 1, 2024, to November 2, 2024)."
#| fig-subcap: ["Unweighted","Weighted"]
#| layout-ncol: 1
#| label: fig-support-trend
# Plotting with the filtered data
plan(multisession)
##unweighted
ggplot(combined_data_total,aes(x = end_date,y=pct, colour = candidate)) +
  geom_point(alpha = 0.1)+ 
  geom_smooth(data = combined_predictions_filtered,aes(x=end_date,y = predicted_unweighted), method = "lm",size = 1)+ 
  geom_line(data = combined_predictions_filtered,aes(x=end_date,y = predicted_unweighted),size = 0.2)+
  labs(
       x = "End Date of the Poll",  # Change x-axis label to End Date
       y = "Predicted Support (%)",
       color = "Candidate",caption = 
                   "Model: Candidate Support Percentage = national_poll + pollster + population + state") +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1),strip.background = element_rect(colour = "black", fill = "white")) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date)) +
  annotate("text", x = max(combined_predictions_filtered$end_date) -0.5, y = 48, 
           label = "47.5%", color = "blue", hjust = -0.2, size = 2) +  # Adjust size here
  annotate("text", x = max(combined_predictions_filtered$end_date) - 0.5, y = 46, 
           label = "46.5%", color = "red", hjust = -0.2, size = 2)

##weighted
ggplot(combined_data_total,aes(x = end_date,y=pct, colour = candidate,size = combined_weight)) +
  geom_point(alpha = 0.1)+ 
  geom_smooth(data = combined_predictions_filtered,aes(x=end_date,y = predicted_unweighted), method = "lm",size = 1)+ 
  geom_line(data = combined_predictions_filtered,aes(x=end_date,y = predicted_weighted), size = 0.2)+
  labs(x = "End Date of the Poll",
       y = "Predicted Support (%)",
       color = "Candidate",
       size="Weight",
       caption = "Weights: recency_weight * sample_size_weight \nModel: Candidate Support Percentage = national_poll + pollster + population + state") +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red"))  +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1),strip.background = element_rect(colour = "black", fill = "white")) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date)) +
  annotate("text", x = max(combined_predictions_filtered$end_date) -0.5, y = 48, 
           label = "47.5%", color = "blue", hjust = -0.2, size = 2) +  # Adjust size here
  annotate("text", x = max(combined_predictions_filtered$end_date) - 0.5, y = 46, 
           label = "46.5%", color = "red", hjust = -0.2, size = 2)
```

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| fig-subcap: ["Unweighted","Weighted"]
#| layout-ncol: 1
#| fig-cap: "Harris is likely to lead Trump in Pennsylvania, Michigan, and Wisconsin by Model Prediction (From August 1, 2024, to November 2, 2024)."
#| label: fig-support-state
## unweighted
plan(multisession)
selected_states <- c("Arizona", "Georgia", "Michigan", "Nevada", "North Carolina", "Pennsylvania", "Wisconsin")
combined_data_total<-combined_data_total|>filter(state %in% selected_states)
combined_predictions_filtered<-combined_predictions_filtered|>filter(state %in% selected_states)
late<-combined_predictions_filtered[which.max(combined_predictions_filtered$end_date), ]
ggplot(combined_data_total,aes(x = end_date,y=pct, colour = candidate))+
  geom_point(alpha = 0.1)+ 
  geom_smooth(data = combined_predictions_filtered,aes(x=end_date,y = predicted_unweighted),se = FALSE, size = 0.5)+
  facet_wrap(~ state,ncol = 4) +
  labs(
       x = "End Date of the Poll",  # Change x-axis label to End Date
       y = "Predicted Support (%)",
       color = "Candidate",
       caption = "Model: Candidate Support Percentage = national_poll + pollster + population + state") +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, hjust = 1, size =6),
        legend.box = "vertical",
        strip.background = element_rect(colour = "black", fill = "white")) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date))

## weighetd
ggplot(combined_data_total,aes(x = end_date,y=pct, colour = candidate,size = combined_weight)) +
  geom_point(alpha = 0.1)+ 
  geom_smooth(data = combined_predictions_filtered,aes(x=end_date,y = predicted_weighted),se = FALSE, size = 0.5)+ # Smoothed trend line
  facet_wrap(~ state,ncol = 4) +
  labs(
       x = "End Date of the Poll",  # Change x-axis label to End Date
       y = "Predicted Support (%)",
       color = "Candidate",
       size="Weight",
       caption = "Weights: recency_weight * sample_size_weight\nModel: Candidate Support Percentage = national_poll + pollster + population + state") +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, hjust = 1,size=6),
        legend.box = "vertical",
        strip.background = element_rect(colour = "black", fill = "white")) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date)) 
```

@@

ggplot(combined_data_total, aes(x = end_date, y = pct)) +
  geom_point(aes(fill = pollster_group), alpha = 0.7, shape = 21) + 
  geom_smooth(data = combined_predictions_filtered %>% filter(candidate == "Kamala Harris"), 
            aes(x = end_date, y = predicted_unweighted), color = "blue", size = 1) +
  geom_smooth(data = combined_predictions_filtered %>% filter(candidate == "Donald Trump"), 
            aes(x = end_date, y = predicted_unweighted), color = "red", size = 1) +
  labs(
    x = "End Date of the Poll",
    y = "Predicted Support (%)",
    color = "Candidate",
    fill = "Top 5 Pollster",
    caption = "Model: Candidate Support Percentage = national_poll + pollster + population + state"
  )+ # Smoothed trend line
  facet_wrap(~ candidate,ncol = 2) +
  scale_fill_manual(values = c(setNames(rep("grey", length(unique(combined_data_total$pollster_group))), "Other"),
                               setNames(brewer.pal(10, "Set1"), top_pollsters))) + 
  theme_classic() +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",  # Place legends in a horizontal row
    legend.title.align = 0.5,   # Center-align legend titles for better appearance
    axis.text.x = element_text(angle = 45, hjust = 1,size=8),
    strip.background = element_rect(colour = "black", fill = "white")
  ) +
  guides(
    size = guide_legend(order = 1, title.position = "top", label.position = "left"),
    fill = guide_legend(order = 2)
  ) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date))

ggplot(combined_data_total, aes(x = end_date, y = pct, size = combined_weight)) +
  geom_point(aes(fill = pollster_group), alpha = 0.7, shape = 21) + 
  geom_smooth(data = combined_predictions_filtered %>% filter(candidate == "Kamala Harris"), 
            aes(x = end_date, y = predicted_weighted), color = "blue", size = 1) +
  geom_smooth(data = combined_predictions_filtered %>% filter(candidate == "Donald Trump"), 
            aes(x = end_date, y = predicted_weighted), color = "red", size = 1)+
  labs(
    x = "End Date of the Poll",
    y = "Predicted Support (%)",
    size = "Weight",
    fill = "Top 5 Pollster",
    caption = "Weights: recency_weight * sample_size_weight \nModel: Candidate Support Percentage = national_poll + pollster + population + state"
  )+ # Smoothed trend line
  facet_wrap(~ candidate,ncol = 2) +
  scale_fill_manual(values = c(setNames(rep("grey", length(unique(combined_data_total$pollster_group))), "Other"),
                               setNames(brewer.pal(10, "Set1"), top_pollsters))) +
  theme_classic() +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",  # Place legends in a horizontal row
    legend.title.align = 0.5,   # Center-align legend titles for better appearance
    axis.text.x = element_text(angle = 45, hjust = 1,size=8),
    strip.background = element_rect(colour = "black", fill = "white")
  ) +
  guides(
    size = guide_legend(order = 1, title.position = "top", label.position = "left"),
    fill = guide_legend(order = 2)
  ) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date))

>>>>

ggplot(harris_data, aes(x = end_date, y = pct, color = factor(numeric_grade))) +
  geom_point(alpha = 0.7) +
  geom_smooth(se = FALSE, span = 0.3) +
  labs(y = "Support for Harris (%)", x = "End Date of the Poll", color = "Numeric Grade") +
  scale_color_viridis_d() +
  theme_classic() +
  theme(legend.position = "bottom") +
  coord_cartesian(ylim = c(0, 100))
```
@tbl-harrissupport shows Harris' average support across states in the 2024 U.S. Presidential Election. She leads in Democratic-leaning states, with top support in Vermont (68.5%), Maryland (62.9%), and Massachusetts (61.2%). In key battlegrounds, Harris's support hovers around 50%, including Wisconsin (49.0%) and Pennsylvania (48.3%). Nationally, she averages 47.9%, with competitive support in swing states like Nevada (47.9%) and North Carolina (47.8%). However, in Republican-leaning states like Texas (44.3%) and Florida (44.2%), her support is lower. This distribution highlights Harris’s strength in Democratic regions and mixed prospects in competitive areas.
```{r}
#| label: tbl-harrissupport
#| tbl-cap: Harris leads in Democratic-leaning states, showing strong support in reliably blue areas. In pivotal swing states, she maintains competitive support in Wisconsin and Pennsylvania, reflecting her appeal in regions important for a potential electoral victory. However, in traditionally Republican strongholds, her support is lower. Nationally, Harris averages 47.9%.
#| echo: false
#| warning: false
#| message: false

@@

# Display the result as a table
kable(harris_data_state)

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| fig-cap: "Harris Kamala leads Donald Trump an average of 1% in Predicted Support in the 2024 US Election (From August 1, 2024, to November 2, 2024)."
#| fig-subcap: ["Unweighted","Weighted"]
#| layout-ncol: 1
#| label: fig-support-trend
# Plotting with the filtered data
plan(multisession)
##unweighted
ggplot(combined_data_total,aes(x = end_date,y=pct, colour = candidate)) +
  geom_point(alpha = 0.1)+ 
  geom_smooth(data = combined_predictions_filtered,aes(x=end_date,y = predicted_unweighted), method = "lm",size = 1)+ 
  geom_line(data = combined_predictions_filtered,aes(x=end_date,y = predicted_unweighted),size = 0.2)+
  labs(
       x = "End Date of the Poll",  # Change x-axis label to End Date
       y = "Predicted Support (%)",
       color = "Candidate",caption = 
                   "Points are polls.\nModel: Candidate Support Percentage = national_poll + pollster + population + state\nLatetst Poll: Nov 2\nSource: FiftyThirtyEight") +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1),strip.background = element_rect(colour = "black", fill = "white")) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date)) +
  annotate("text", x = max(combined_predictions_filtered$end_date) -0.5, y = 48, 
           label = "47.5%", color = "blue", hjust = -0.2, size = 2) +  # Adjust size here
  annotate("text", x = max(combined_predictions_filtered$end_date) - 0.5, y = 46, 
           label = "46.5%", color = "red", hjust = -0.2, size = 2)

##weighted
ggplot(combined_data_total,aes(x = end_date,y=pct, colour = candidate,size = combined_weight)) +
  geom_point(alpha = 0.1)+ 
  geom_smooth(data = combined_predictions_filtered,aes(x=end_date,y = predicted_unweighted), method = "lm",size = 1)+ 
  geom_line(data = combined_predictions_filtered,aes(x=end_date,y = predicted_weighted), size = 0.2)+
  labs(x = "End Date of the Poll",
       y = "Predicted Support (%)",
       color = "Candidate",
       size="Weight",
       caption = "Lines are weighted by recency_weight * sample_size_weight.\nPoints are polls.\nModel: Candidate Support Percentage = national_poll + pollster + population + state\nLatetst Poll: Nov 2\nSource: FiftyThirtyEight") +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red"))  +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1),strip.background = element_rect(colour = "black", fill = "white")) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date)) +
  annotate("text", x = max(combined_predictions_filtered$end_date) -0.5, y = 48, 
           label = "47.5%", color = "blue", hjust = -0.2, size = 2) +  # Adjust size here
  annotate("text", x = max(combined_predictions_filtered$end_date) - 0.5, y = 46, 
           label = "46.5%", color = "red", hjust = -0.2, size = 2)
```

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| fig-subcap: ["Unweighted","Weighted"]
#| layout-ncol: 1
#| fig-cap: "Harris is likely to lead Trump in Pennsylvania, Michigan, and Wisconsin by Model Prediction (From August 1, 2024, to November 2, 2024)."
#| label: fig-support-state
## unweighted
plan(multisession)
selected_states <- c("Arizona", "Georgia", "Michigan", "Nevada", "North Carolina", "Pennsylvania", "Wisconsin")
combined_data_total<-combined_data_total|>filter(state %in% selected_states)
combined_predictions_filtered<-combined_predictions_filtered|>filter(state %in% selected_states)
late<-combined_predictions_filtered[which.max(combined_predictions_filtered$end_date), ]
ggplot(combined_data_total,aes(x = end_date,y=pct, colour = candidate))+
  geom_point(alpha = 0.1)+ 
  geom_smooth(data = combined_predictions_filtered,aes(x=end_date,y = predicted_unweighted),se = FALSE, size = 0.5)+
  facet_wrap(~ state,ncol = 4) +
  labs(
       x = "End Date of the Poll",  # Change x-axis label to End Date
       y = "Predicted Support (%)",
       color = "Candidate",
       caption = "Points are polls.\nModel: Candidate Support Percentage = national_poll + pollster + population + state\nLatetst Poll: Nov 2\nSource: FiftyThirtyEight") +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, hjust = 1, size =6),
        legend.box = "vertical",
        strip.background = element_rect(colour = "black", fill = "white")) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date))

## weighetd
ggplot(combined_data_total,aes(x = end_date,y=pct, colour = candidate,size = combined_weight)) +
  geom_point(alpha = 0.1)+ 
  geom_smooth(data = combined_predictions_filtered,aes(x=end_date,y = predicted_weighted),se = FALSE, size = 0.5)+ # Smoothed trend line
  facet_wrap(~ state,ncol = 4) +
  labs(
       x = "End Date of the Poll",  # Change x-axis label to End Date
       y = "Predicted Support (%)",
       color = "Candidate",
       size="Weight",
       caption = "Lines are weighted by recency_weight * sample_size_weight.\nPoints are polls.\nModel: Candidate Support Percentage = national_poll + pollster + population + state\nLatetst Poll: Nov 2\nSource: FiftyThirtyEight") +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, hjust = 1,size=6),
        legend.box = "vertical",
        strip.background = element_rect(colour = "black", fill = "white")) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date)) 
```

@@

ggplot(combined_data_total, aes(x = end_date, y = pct)) +
  geom_point(aes(fill = pollster_group), alpha = 0.7, shape = 21) + 
  geom_smooth(data = combined_predictions_filtered %>% filter(candidate == "Kamala Harris"), 
            aes(x = end_date, y = predicted_unweighted), color = "blue", size = 1) +
  geom_smooth(data = combined_predictions_filtered %>% filter(candidate == "Donald Trump"), 
            aes(x = end_date, y = predicted_unweighted), color = "red", size = 1) +
  labs(
    x = "End Date of the Poll",
    y = "Predicted Support (%)",
    color = "Candidate",
    fill = "Top 5 Pollster",
    caption = "Points are polls.\nModel: Candidate Support Percentage = national_poll + pollster + population + state\nLatetst Poll: Nov 2\nSource: FiftyThirtyEight"
  )+ # Smoothed trend line
  facet_wrap(~ candidate,ncol = 2) +
  scale_fill_manual(values = c(setNames(rep("grey", length(unique(combined_data_total$pollster_group))), "Other"),
                               setNames(brewer.pal(10, "Set1"), top_pollsters))) + 
  theme_classic() +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",  # Place legends in a horizontal row
    legend.title.align = 0.5,   # Center-align legend titles for better appearance
    axis.text.x = element_text(angle = 45, hjust = 1,size=8),
    strip.background = element_rect(colour = "black", fill = "white")
  ) +
  guides(
    size = guide_legend(order = 1, title.position = "top", label.position = "left"),
    fill = guide_legend(order = 2)
  ) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date))

ggplot(combined_data_total, aes(x = end_date, y = pct, size = combined_weight)) +
  geom_point(aes(fill = pollster_group), alpha = 0.7, shape = 21) + 
  geom_smooth(data = combined_predictions_filtered %>% filter(candidate == "Kamala Harris"), 
            aes(x = end_date, y = predicted_weighted), color = "blue", size = 1) +
  geom_smooth(data = combined_predictions_filtered %>% filter(candidate == "Donald Trump"), 
            aes(x = end_date, y = predicted_weighted), color = "red", size = 1)+
  labs(
    x = "End Date of the Poll",
    y = "Predicted Support (%)",
    size = "Weight",
    fill = "Top 5 Pollster",
    caption = "Lines are weighted by recency_weight * sample_size_weight.\nPoints are polls.\nModel: Candidate Support Percentage = national_poll + pollster + population + state\nLatetst Poll: Nov 2\nSource: FiftyThirtyEight"
  )+ # Smoothed trend line
  facet_wrap(~ candidate,ncol = 2) +
  scale_fill_manual(values = c(setNames(rep("grey", length(unique(combined_data_total$pollster_group))), "Other"),
                               setNames(brewer.pal(10, "Set1"), top_pollsters))) +
  theme_classic() +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",  # Place legends in a horizontal row
    legend.title.align = 0.5,   # Center-align legend titles for better appearance
    axis.text.x = element_text(angle = 45, hjust = 1,size=8),
    strip.background = element_rect(colour = "black", fill = "white")
  ) +
  guides(
    size = guide_legend(order = 1, title.position = "top", label.position = "left"),
    fill = guide_legend(order = 2)
  ) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date))
